---
title: "Building a Survey Questionnaire with qretools"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Building a Survey Questionnaire with qretools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
library(qretools)
```

This vignette shows how to use qretools to build a survey questionnaire from
start to finish.  All examples use the built-in example project shipped with
the package, so you can run every code block without any additional files.

## How qretools organises a project

A qretools project is a directory that contains:

| Path | Purpose |
|---|---|
| `project-config.yml` | Project-level metadata and path overrides |
| `banks/questions/` | One or more YAML files defining question variables |
| `banks/control/` | Control-parameter variables (survey metadata, A/B flags, …) |
| `banks/modules/` | Reusable question modules |
| `value-labels/` | Shared value-label sets |
| `surveys/<id>/design/` | Per-survey YAML configuration files |

`qt_config()` locates the project root automatically by walking up the
directory tree until it finds `project-config.yml` or `_qretools.yml`.

```{r point-at-examples}
# Point the session at the built-in example project
example_dir <- system.file("examples", package = "qretools")
setwd(example_dir)

config <- qt_config()
config
```

## Loading the question bank

Before building a questionnaire you can inspect the banks independently:

```{r load-banks}
vlabs  <- qt_read_value_labels(config)
qbank  <- qt_read_question_bank(config, value_labels = vlabs)
qbank
```

Each entry in the question bank is a `qt_qvar` object.  Print one to see its
fields:

```{r inspect-variable}
qbank$variables[["nhd_sat"]]
```

## Two ways to build a questionnaire

qretools supports two authoring styles that produce identical output.

### Option A — write the YAML by hand

A survey config file has two top-level keys: `meta` (survey metadata) and
`questionnaire` (the ordered item list).

```yaml
meta:
  id: bas-2026
  title: "Baltimore Area Survey 2026"
  status: draft
  controls_required: [YEAR, JURISDICTION]

questionnaire:
  items:
    - item_type: section
      id: neighborhoods
      title: Neighborhoods
      items:
        - item_type: question
          variable_id: nhd_sat
```

Read and expand a YAML file with `qt_read_survey_config()`:

```{r read-yaml, eval=FALSE}
qreconfig <- qt_read_survey_config(
  "surveys/bas-example/design/survey-bas-example.yml"
)
```

### Option B — build programmatically with `qt_qre()`

The builder mirrors the YAML structure but lets you compose questionnaires
in R scripts and pipe chains.  Every `qt_add_*()` function works in two modes:

- **Pipe mode** — first argument is a `qt_qre` object; the item is appended and
  the modified `qt_qre` is returned invisibly.
- **Spec mode** — first argument is an ID string; the function returns a plain
  item spec for nesting inside a container such as `qt_add_section()`.

```{r build-programmatic, eval=FALSE}
qre <- qt_qre(
  id                = "bas-2026",
  title             = "Baltimore Area Survey 2026",
  status            = "draft",
  organization      = "Johns Hopkins 21st Century Cities Initiative",
  controls_required = c("YEAR", "JURISDICTION")
)
```

## Adding sections and questions

Questions are referenced by their `variable_id` in the question bank.
Sections group related questions:

```{r add-sections, eval=FALSE}
qre <- qre |>
  qt_add_section("neighborhoods", "Neighborhoods",
    qt_add_question("nhd_sat"),
    qt_add_question("nhd_chg")
  ) |>
  qt_add_section("demographics", "About You",
    qt_add_question("dem_yearborn"),
    qt_add_question("dem_race"),
    qt_add_question("dem_edattain")
  )
```

## Skip patterns with `qt_add_logic()`

A logic block routes survey flow based on an R expression.  Items in `then`
are shown when the condition is `TRUE`; items in `otherwise` are shown when
it is `FALSE`.  The `otherwise` argument serializes as `else:` in YAML.

```{r skip-patterns, eval=FALSE}
qre <- qre |>
  qt_add_section("housing", "Housing",

    # Ask about tenure first
    qt_add_question("hsg_tenure"),

    # Skip to the relevant follow-up based on tenure
    qt_add_logic("hsg_owner_logic",
      condition  = "hsg_tenure == 2",        # 2 = Own (see value label set 'own')
      then       = list(
        qt_add_question("hsg_own_years"),
        qt_add_question("hsg_own_value")
      ),
      otherwise  = list(
        qt_add_question("hsg_rent_monthly")
      )
    )
  )
```

Logic blocks can be nested.  For example, a follow-up question about children
only appears after a screening question confirms the respondent has children:

```{r nested-skip, eval=FALSE}
qt_add_section("demographics", "About You",
  qt_add_question("dem_child"),              # "Do you have children at home?"
  qt_add_logic("dem_children_logic",
    condition = "dem_child == 1",
    then = list(
      qt_add_question("dem_children_n"),
      qt_add_question("dem_children_age_youngest")
    )
  )
)
```

## Select-all (multiple-response) questions

There are two ways to represent a "select all that apply" question.

### Approach 1 — separate indicator variables (recommended)

This matches the convention used in the BAS question banks.  Each response
option is its own variable with `response_type: factor` and a yes/no value
label set.  All share the same `question_text`.

**In the question bank (`banks/questions/demographics.yml`):**

```yaml
dem_race_wh:
  title: "Race indicator: White"
  response_type: factor
  question_text: "What is your racial or ethnic background? Select all that apply."
  surveys_used: [bas-2026]
  vargroup: dem
  value_label_id: yesno01

dem_race_bl:
  title: "Race indicator: Black or African American"
  response_type: factor
  question_text: "What is your racial or ethnic background? Select all that apply."
  surveys_used: [bas-2026]
  vargroup: dem
  value_label_id: yesno01

dem_race_hisp:
  title: "Race indicator: Hispanic or Latino"
  response_type: factor
  question_text: "What is your racial or ethnic background? Select all that apply."
  surveys_used: [bas-2026]
  vargroup: dem
  value_label_id: yesno01
```

**In the questionnaire,** group them with `qt_add_display_together()` so the
survey platform knows they belong to one screen:

```{r select-all-indicators, eval=FALSE}
qt_add_section("demographics", "About You",
  qt_add_display_together("dem_race_block",
    qt_add_question("dem_race_wh",   hide_question = TRUE),
    qt_add_question("dem_race_bl",   hide_question = TRUE),
    qt_add_question("dem_race_hisp", hide_question = TRUE)
  )
)
```

`hide_question = TRUE` suppresses the repeated question text on the second
and subsequent items, showing it only once for the whole block.

### Approach 2 — `select_all` response type

Use this when you want a single question bank entry to document the whole
select-all item.  The entry must list the indicator variables it creates
(`creates_variables`) and define each option (`variable_parts`).

Each option has two label fields:

- **`option_text`** — the text the respondent sees on screen ("White").
- **`option_title`** — the subordinate title appended to the parent `title`
  for codebook documentation.  Composed as `"{title}: {option_title}"`, so
  `title: "Race indicator"` + `option_title: "White"` → codebook entry
  `"Race indicator: White"`.  When omitted, defaults to `option_text`.

When `option_title` and `option_text` are the same, you can use a simple
string shorthand (the string is treated as `option_text`, with `option_title`
defaulting to the same value):

```yaml
dem_race_origin:
  title: Race indicator
  response_type: select_all
  question_text: "What is your racial or ethnic background? Select all that apply."
  surveys_used: [bas-2026]
  vargroup: dem
  creates_variables: [dem_race_wh, dem_race_bl, dem_race_hisp, dem_race_as, dem_race_oth]
  variable_parts:
    dem_race_wh:   White                        # shorthand: option_text = option_title = "White"
    dem_race_bl:   Black or African American    # → codebook: "Race indicator: Black or African American"
    dem_race_hisp: Hispanic or Latino
    dem_race_as:   Asian or Pacific Islander
    dem_race_oth:  Some other race or ethnicity
```

When the respondent-facing text and codebook label differ, use the full object
form:

```yaml
dem_race_origin:
  title: Race indicator
  response_type: select_all
  question_text: "What is your racial or ethnic background? Select all that apply."
  surveys_used: [bas-2026]
  vargroup: dem
  creates_variables: [dem_race_wh, dem_race_bl, dem_race_hisp]
  variable_parts:
    dem_race_wh:
      option_title: White                  # → codebook: "Race indicator: White"
      option_text:  White
    dem_race_bl:
      option_title: Black or African American
      option_text:  Black or African American
    dem_race_hisp:
      option_title: Hispanic or Latino
      option_text:  "Hispanic, Latino, or Spanish origin"   # longer phrasing on screen
```

In the questionnaire this is a single question reference:

```{r select-all-mrtype, eval=FALSE}
qt_add_question("dem_race_origin")
```

### Other/Specify options

When one of the select-all options is "Other — please specify", it produces
**two** variables:

1. **The indicator** — a boolean variable (like every other select-all part)
   that records whether the respondent chose the "Other" option at all.
2. **The specify text variable** — a companion character variable that stores
   whatever the respondent typed.

Both are documented inside the same `variable_parts` entry by adding an
`other_specify` sub-object.  Use `other_title` to give the text variable its
own codebook title, distinct from the indicator's `option_title`:

```yaml
dem_race_origin:
  title: Race indicator
  response_type: select_all
  question_text: "What is your racial or ethnic background? Select all that apply."
  surveys_used: [bas-2026]
  vargroup: dem
  creates_variables: [dem_race_wh, dem_race_bl, dem_race_hisp, dem_race_as, dem_race_oth]
  variable_parts:
    dem_race_wh:   White
    dem_race_bl:   Black or African American
    dem_race_hisp: Hispanic or Latino
    dem_race_as:   Asian or Pacific Islander
    dem_race_oth:
      option_title: Other race or ethnicity          # title for dem_race_oth (boolean)
      option_text:  Some other race or ethnicity     # text shown on screen
      other_specify:
        id:               dem_race_oth_text          # creates second variable
        other_title:      "Other race (specify)"     # title for dem_race_oth_text
        other_label:      Please specify             # label next to the text box on screen
        character_length: 200
```

This produces two variables in the bank:

| Variable | Storage | Codebook title |
|---|---|---|
| `dem_race_oth` | boolean | Race indicator: Other race or ethnicity |
| `dem_race_oth_text` | character | Race indicator: Other race (specify) |

If `other_title` is omitted, the text variable's codebook title falls back to
`other_label` (or "Specify" if that is also absent).

## Preload computes

Compute items in the `preload` block run before any questions are displayed —
useful for assigning experimental conditions or deriving values from control
parameters:

```{r preload, eval=FALSE}
qre <- qre |>
  qt_add_preload(
    qt_add_compute(
      id          = "assign_condition",
      description = "Assign respondent to A/B experimental condition",
      command     = "condition <- sample(c('A', 'B'), 1)",
      provides    = "condition"
    ),
    qt_add_compute(
      id          = "set_jurisdiction_label",
      description = "Set short label for jurisdiction",
      command     = "jlabel <- if (JURISDICTION == '1') 'City' else 'County'",
      requires    = "JURISDICTION",
      provides    = "jlabel"
    )
  )
```

A split block then routes respondents based on a computed value:

```{r split, eval=FALSE}
qt_add_section("experiment", "Community Questions",
  qt_add_split("community_wording",
    description = "A/B wording test",
    control_id  = "condition",
    paths = list(
      qt_add_path("wording_a", control_value = "A",
        qt_add_question("com_attach_a")),
      qt_add_path("wording_b", control_value = "B",
        qt_add_question("com_attach_b"))
    )
  )
)
```

## Loops

A loop repeats a set of items, substituting `{i}` in variable IDs and titles
with values from `id_fills`.  The question bank must have a **template entry**
whose key contains the literal string `{i}`:

```yaml
# banks/questions/neighborhoods.yml
"feat_rating_{i}":
  title: "Rating of {i}"
  response_type: factor
  question_text: "How would you rate {i} in your neighborhood?"
  surveys_used: [bas-2026]
  vargroup: nhd
  value_label_id: excellent5
```

In the questionnaire:

```{r loop, eval=FALSE}
qt_add_section("features", "Neighborhood Features",
  qt_add_statement("features_intro",
    text = "We will ask you about a few features of your neighborhood."),
  qt_add_loop("feature_ratings",
    id_fills    = c("1", "2", "3"),
    title_fills = c("parks and green space", "public safety", "local businesses"),
    qt_add_question("feat_rating_{i}"),
    qt_add_question("feat_improve_{i}")
  )
)
```

The expander looks up `feat_rating_{i}` in the bank for every iteration and
then substitutes `{i}` in the final variable ID (producing `feat_rating_1`,
`feat_rating_2`, `feat_rating_3`) and in the title.

## Processing and inspecting the result

### From a `qt_qre` object

`qt_process()` expands all containers, resolves every question reference
against the bank, and returns a `qt_qreconfig` object.  It also writes the
YAML to disk so you have a version-controllable artifact:

```{r process, eval=FALSE}
qreconfig <- qt_process(qre)
# Survey YAML written to: surveys/bas-2026/design/survey-bas-2026.yml
```

To suppress the write step, pass `yaml_path = NULL`:

```{r process-no-write, eval=FALSE}
qreconfig <- qt_process(qre, yaml_path = NULL)
```

### From an existing YAML file

```{r process-yaml}
qreconfig <- qt_process(
  "surveys/bas-example/design/survey-bas-example.yml"
)
qreconfig
```

### What you get back

The `qt_qreconfig` object contains:

```{r inspect}
# Total number of expanded items (questions, statements, computes)
length(qreconfig$items)

# Variable IDs of all question items
questions <- Filter(function(i) i$item_type == "question", qreconfig$items)
vapply(questions, `[[`, character(1), "variable_id")

# Questions grouped by section
str(qreconfig$indices$section_map, max.level = 1)
```

Each question item carries its full definition from the bank alongside
provenance fields (section, loop iteration, logic branch, etc.) that make
it straightforward to reconstruct the survey flow programmatically:

```{r inspect-item}
q <- qreconfig$items[[which(
  vapply(qreconfig$items, function(i) identical(i$variable_id, "nhd_sat"),
         logical(1))
)]]

cat("Variable: ", q$variable_id, "\n")
cat("Section:  ", q$section_id,  "\n")
cat("Title:    ", q$definition$title, "\n")
cat("Type:     ", q$definition$storage_type, "\n")
```
