---
title: "Getting Started with qretools"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with qretools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>"
)
library(qretools)
```

`qretools` helps you manage survey questions, generated variables, and
control parameters across multiple survey waves. This vignette covers:

1. Setting up a project
2. Loading the variable banks
3. Querying banks for variable documentation and search

---

## 1. Project setup

### Directory structure

A `qretools` project lives in a single root directory. Inside that directory,
the following layout is expected by default:

```
my-survey/
├── project-config.yml       # Required: project metadata
├── _qretools.yml            # Optional: override default paths
├── banks/
│   ├── questions/           # Question bank (one or more .yml files)
│   ├── generated/           # Generated variable bank
│   └── control/             # Control parameter bank
└── value-labels/            # Value label definitions
    └── value-labels.yml
```

Each bank directory can contain a single `.yml` file or multiple `.yml` files
(one per topic or questionnaire module). `qretools` reads and combines all
`.yml` files it finds in a directory.

### Required file: `project-config.yml`

The only required file is `project-config.yml` at the project root. At minimum
it must contain three fields:

```yaml
project:
  title: "Baltimore Area Survey"
  organization: "American University"
  principal_investigator: "Michael Bader"

irb:
  institution: "American University"
  protocol_number: "IRB-2020-001"
```

`principal_investigator` can be a plain string (as above) or a structured
record with `name`, `affiliation`, and `email` sub-fields:

```yaml
project:
  title: "Baltimore Area Survey"
  organization: "American University"
  principal_investigator:
    name: "Michael Bader"
    affiliation: "American University"
    email: "mbader@american.edu"
```

For projects with multiple PIs, supply a list of such records.

### Optional file: `_qretools.yml`

If your project uses non-default directory names, add a `_qretools.yml` file
at the project root to override any path. You only need to list paths that
differ from the defaults:

```yaml
paths:
  question_bank: "instruments/questions"
  value_labels:  "instruments/value-labels"
```

All other paths default to the standard layout shown above.

### Loading the configuration

Point `qt_config()` at the project root. When called from inside the project
directory, the `project_root` argument can be omitted and the current working
directory is used:

```{r load-config-user, eval=FALSE}
library(qretools)

# From the project root directory (most common during analysis)
config <- qt_config()

# Or with an explicit path
config <- qt_config(project_root = "~/projects/my-survey")
```

For this vignette we use the example project bundled with the package:

```{r load-config-example}
example_root <- system.file("examples", package = "qretools")
config <- qt_config(project_root = example_root)
```

### Verifying the configuration

`print()` gives a one-line confirmation that the configuration loaded and
shows the project root:

```{r print-config}
print(config)
```

`summary()` shows full detail — project metadata, which configuration files
were found, the resolved paths for each bank, and cache status. This is the
first thing to check when something isn't loading as expected:

```{r summary-config}
summary(config)
```

If `_qretools.yml` was found, its path appears with a checkmark under
**Configuration Files**. The **Key Paths** section shows where `qretools`
will look for each bank relative to the project root.

### Performance: caching

For large projects, `qt_config()` supports two caching modes so that
configuration files are not re-parsed on every call:

```{r caching, eval=FALSE}
# Cache in memory for the duration of the R session
config <- qt_config(cache = "session")

# Cache to disk so it persists across R sessions
config <- qt_config(cache = "persist")

# Clear the cache manually (rarely needed — cache auto-invalidates
# when project-config.yml or _qretools.yml changes on disk)
qt_clear_cache()
```

---

## 2. Loading variable banks

### Value labels

Load value labels first. They are used to validate factor variables in all
other banks and to resolve labels in codebook output:

```{r load-vlabs}
vl <- qt_read_value_labels(config)
print(vl)
```

The shorter alias `qt_vlabs()` is equivalent:

```{r load-vlabs-alias, eval=FALSE}
vl <- qt_vlabs(config)
```

Individual label sets are stored in `vl$labels` and can be inspected
directly:

```{r inspect-vlabs}
vl$labels[["satisfied5"]]
```

### Question bank

```{r load-qbank}
qbank <- qt_read_question_bank(config)
print(qbank)
```

`qt_qbank()` is the short alias:

```{r load-qbank-alias, eval=FALSE}
qbank <- qt_qbank(config)
```

The printed summary shows how many variables are in the bank and which survey
waves they span.

### Generated variable bank and control parameter bank

The generated variable and control parameter banks are loaded the same way.
Your project will need corresponding `banks/generated/` and `banks/control/`
directories (or alternative paths configured in `_qretools.yml`):

```{r load-other-banks, eval=FALSE}
genbank  <- qt_read_generated_variables(config)
ctrlbank <- qt_read_control_parameters(config)

# Short aliases
genbank  <- qt_genbank(config)
ctrlbank <- qt_ctrlbank(config)
```

### What the bank object contains

Every bank object has the same two-part structure:

| Component | Description |
|---|---|
| `$variables` | Named list of variable objects, indexed by variable ID |
| `$meta` | Metadata: source path, file list, read time, and indices |

```{r bank-structure}
# Total variable count
qbank$meta$n_variables

# Source location
qbank$meta$source_path

# Files that were read
qbank$meta$source_files
```

---

## 3. Querying banks

### Accessing a specific variable

Use `$variables` with the variable ID to retrieve a single variable:

```{r access-var}
nhd_sat <- qbank$variables[["nhd_sat"]]
```

`print()` gives the brief format — variable ID, storage type, title, and
the survey waves in which the variable was fielded:

```{r print-var}
print(nhd_sat)
```

To inspect the full contents of the underlying list, use `str()` or access
fields directly:

```{r inspect-var}
nhd_sat$title
nhd_sat$question_text
nhd_sat$storage_type
nhd_sat$surveys_used
nhd_sat$vargroup
nhd_sat$value_label_id
nhd_sat$source
```

### Available indices

Each bank carries four pre-built indices in `$meta$indices`:

| Index | Description |
|---|---|
| `by_variable_id` | Character vector of all IDs, sorted alphabetically |
| `by_file_position` | Character vector of IDs in the order they appear in the source files |
| `by_file` | Named list: each element is a character vector of variable IDs from one file |
| `by_survey` | Named list: each element is a character vector of variable IDs used in one survey wave |

```{r show-indices}
# Alphabetical list of all variable IDs
qbank$meta$indices$by_variable_id

# Variables grouped by source file
qbank$meta$indices$by_file

# Variables grouped by survey wave
qbank$meta$indices$by_survey
```

### Checking whether a variable was fielded in a given survey

The `by_survey` index maps survey IDs to the variables used in each wave.
The most direct check is:

```{r by-survey-check}
# Was nhd_sat asked in bas-2023?
"nhd_sat" %in% qbank$meta$indices$by_survey[["bas-2023"]]

# Was nhd_blks asked in bas-2023?
"nhd_blks" %in% qbank$meta$indices$by_survey[["bas-2023"]]
```

To retrieve all variables fielded in a specific wave:

```{r by-survey-list}
# All variable IDs used in bas-2024
qbank$meta$indices$by_survey[["bas-2024"]]
```

To turn those IDs into a named list of variable objects:

```{r by-survey-objects}
ids_2024  <- qbank$meta$indices$by_survey[["bas-2024"]]
vars_2024 <- qbank$variables[ids_2024]

# Print a brief summary of each
lapply(vars_2024, print)
```

To see which survey waves are represented in the bank at all:

```{r all-surveys}
names(qbank$meta$indices$by_survey)
```

### Finding variables by topic group (`vargroup`)

`vargroup` is a short code that groups variables by topic (e.g. `"nhd"` for
neighborhood questions, `"dem"` for demographics). Because no dedicated
`by_vargroup` index is pre-built, filter the `$variables` list directly:

```{r by-vargroup}
nhd_vars <- Filter(
  function(v) identical(v$vargroup, "nhd"),
  qbank$variables
)

# Print each variable in the group
lapply(nhd_vars, print)
```

To get them in the order they appear in the source files (rather than
whatever order `Filter()` returns), use `by_file_position` to control the
sequence:

```{r by-vargroup-ordered}
ordered_ids <- qbank$meta$indices$by_file_position
nhd_ids     <- ordered_ids[ordered_ids %in% names(nhd_vars)]

lapply(qbank$variables[nhd_ids], print)
```

### Getting all variables from a single file (module)

The `by_file` index groups variable IDs by the `.yml` file they came from.
This is useful when files are organized by questionnaire module:

```{r by-file}
# Which files are in this bank?
names(qbank$meta$indices$by_file)

# All variables in the neighborhoods file, in source order
nhd_file_ids <- qbank$meta$indices$by_file[["neighborhoods.yml"]]
nhd_file_ids

lapply(qbank$variables[nhd_file_ids], print)
```

### Getting all variables in file order

`by_file_position` gives a single flat vector of all variable IDs in the
order they appear across all source files. This reproduces the original
questionnaire sequence:

```{r by-file-position}
qbank$meta$indices$by_file_position
```

Use it to iterate over the full bank in order:

```{r by-file-position-iterate}
for (id in qbank$meta$indices$by_file_position) {
  print(qbank$variables[[id]])
}
```

### Finding all surveys that include a variable

Invert the `by_survey` index to find every wave in which a given variable
appeared:

```{r variable-surveys}
# Which waves included nhd_nyrs?
Filter(
  function(survey) "nhd_nyrs" %in% qbank$meta$indices$by_survey[[survey]],
  names(qbank$meta$indices$by_survey)
)
```

Or access `surveys_used` directly from the variable object:

```{r variable-surveys-direct}
qbank$variables[["nhd_nyrs"]]$surveys_used
```

---

## Summary of key functions

| Function | Short alias | Returns |
|---|---|---|
| `qt_config()` | — | `qt_config` object |
| `qt_read_value_labels()` | `qt_vlabs()` | `qt_vlabs` object |
| `qt_read_question_bank()` | `qt_qbank()` | `qt_qbank` object |
| `qt_read_generated_variables()` | `qt_genbank()` | `qt_genbank` object |
| `qt_read_control_parameters()` | `qt_ctrlbank()` | `qt_ctrlbank` object |
| `qt_codebook()` | — | text/HTML/PDF codebook |

| Bank component | What it contains |
|---|---|
| `$variables[["id"]]` | Individual variable object |
| `$meta$n_variables` | Count of variables |
| `$meta$indices$by_variable_id` | Alphabetical ID list |
| `$meta$indices$by_file_position` | IDs in source-file order |
| `$meta$indices$by_file` | IDs grouped by source file |
| `$meta$indices$by_survey` | IDs grouped by survey wave |
